---
title: "eda.rmd"
author: "Matt McAnear"
date: '2023-04-02'
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(data.table)
library(ggplot2)
library(GGally)
library(stringi)
```

```{r}
salaries <- rbindlist(lapply(list.files('../data/salaries', full.names = TRUE), fread))[
    POS != 'G'
]
salaries$PLAYER <- trimws(gsub("^[0-9]+\\.", "", salaries$PLAYER))

clean_name <- function(name) {
    trimws(tolower(gsub('[^A-z]', '', stringi::stri_trans_general(name, "Latin-ASCII"))))
    # trimws(tolower(gsub('[^A-z]', '', iconv(name, "latin1", "ASCII", sub=""))))
}

salaries[, `:=` (
    'SALARY' = as.numeric(gsub('[^0-9]', '', SALARY)),
    'AAV' = as.numeric(gsub('[^0-9]', '', AAV)),
    'BASE SALARY' = as.numeric(gsub('[^0-9]', '', `BASE SALARY`)),
    'S.BONUS' = as.numeric(gsub('[^0-9]', '', `S.BONUS`)),
    'P.BONUS' = as.numeric(gsub('[^0-9]', '', `P.BONUS`)),
    'CAP HIT' = as.numeric(gsub('[^0-9]', '', `CAP HIT`)),
    'CAP HIT %' = as.numeric(gsub('[^0-9]', '', `CAP HIT %`)),
    'SEASON' = as.integer(sapply(strsplit(SEASON, '-'), '[', 1)),
    'SIGNING DATE' = coalesce(
        as.Date(`SIGNING DATE`, format = '%b %d, %Y'), 
        as.Date(`SIGNING DATE`, '%b. %d, %Y')
        ),
    'merge_name' = clean_name(PLAYER)
    )
]
salaries <- salaries[SALARY > 0]

player_stats <- fread('../data/all_player_data.csv')[situation == 'all' & position != 'G']
team_metadata <- fread('../data/team_metadata.csv')

core_stats <- player_stats[,
    .(
        playerId, season, name, position, team, games_played, icetime, shifts, onIce_xGoalsPercentage, 
        offIce_xGoalsPercentage, penalties, faceoffsWon, penalityMinutes, penalityMinutesDrawn, 
        shotsBlockedByPlayer, I_F_goals, I_F_primaryAssists, I_F_secondaryAssists, I_F_points,
        I_F_hits, 'merge_name' = clean_name(name)
     )
]


# core_stats$team <- 
core_stats$team <- sapply(core_stats$team, function(x) {
    new_team <- switch(
        x,
        S.J = 'SJS',
        T.B = 'TBL',
        N.J = "NJD",
        L.A = "LAK"
    )
    ifelse(is.null(new_team), x, new_team)
})
    
all_player_data <- merge(
    core_stats,
    salaries,
    by.y = c('merge_name', 'SEASON'),
    by.x = c('merge_name', 'season'),
    all = FALSE
)
full_df <- merge(
    all_player_data,
    team_metadata,
    by.x = 'team',
    by.y = 'Abbr',
    all = FALSE
)


write.csv(full_df, './data/full_data.csv', row.names = FALSE)

```

The two datasets are not complete - some player's with stats are missing from the salary
data and vice versa. There are ~11,000 unique contracts.

```{r goals-vs-toi}
plot_data <- full_df[
    sample(1:.N, size=1000),
    .(
        'salary' = SALARY,
        'goals' = I_F_goals, 
        'assists' = I_F_primaryAssists, 
        'avg_icetime' = icetime/60/games_played, 
        'points' = I_F_points,
        Division
    )
]
ggpairs(plot_data, aes(col=Division, alpha=0.5), columns=1:5, progress = FALSE) +
    theme_bw()
   
```

No clear difference between divisions and the relationship of salary vs. points/60, on 
either offense or defense.

```{r points-per-60}
ggplot(data = full_df[sample(1:.N, size=2500)], aes(x = SALARY, y = I_F_points/(icetime/3600), col = Division)) +
    geom_point() +
    scale_y_log10() + 
    geom_smooth(method='lm') + 
    scale_color_brewer(palette='Dark2') +
    theme_bw() +
    xlab('Salary') +
    scale_x_log10(labels = scales::dollar_format(scale=1e-6, suffix='M', accuracy = 0.1)) +
    ylab('Pts/60') +
    facet_wrap(~ position) 
```


Many categorical variables, but the ANOVA analysis is overspecified. There is a significant 
relationship with predictable variables (goals and points)


```{r salary-vs-height-weight}
fit <- lm(data=full_df, log(SALARY) ~ `Division` + EffTaxRate + O6 + I_F_goals*position + I_F_hits*position + penalityMinutes*position)
summary(fit)
```


